An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-17 scagnostics.Rmd
Index matrix added for bbh example.

The above example is using only a reduced set of the free parameters relevant to the physical interpretation, and we already know the expected result by construction. In general many additional nuissance parameters will also enter the fit, see e.g. @Smith:2016qas. We use a simulation study \textcolor{red}{simulating a binary black hole (BBH) merger event...need more info maybe, no reference but Rory offered to write short description} including the full set of nuissance parameters as a test case of larger parameter space in which we aim to detect structure. The dataset is summarised in \ref{fig:bbhSimulation} where we have removed a subset of less informative parameters to obtain a clear picture. Interesting correlations can be seen between the parameters "ra", "dec" and "time", where the first two describe the position of the BBH event in the sky, and the "time" being the merging time (in GPS units). Note in particular the relation between "dec" and "time" is not functional, which motivates us to use the TIC index for this example (drawback is however the run time, this data sample contains 10k points, making optimisation of the MINE index functions very inefficient). Some other interesting patterns can also be identified. (These patterns are representative of the likelihood function, since most priors are flat, or build such to capture growth with volume rahter than distance. This is still efficient thanks to special sampling techniques that allow jumps in parameter space.)

As a first test we use the full dataset (with exception of $m2$ for the same reasons as stated in the first example), and use different index functions. Here we expect to recover some relation between the correlated variables dec, time and ra. We compare the final view using three different metrics in Fig. \ref{fig:bbhGuided}, showing the views with (local?) maximum of the TIC index (left) spines2d index (middle) and for comparison maximum of 1-convex (right). Indeed the TIC and spline index identify dec, time and ra parameters and show views of the 2-d band that is embedded in three dimensions \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. Notably, the exact views differ based on the index function selected, optimising TIC is showing us the band "from the side", resulting in clearly non-functional dependence. On the other hand, the index based on the scagnostics member convex is too noisy to find any views of interest, and the final  view is essentially equal to the starting plane in this case. \textcolor{red}{different scagnostics member may be better suited, but I expect the result will be the same. might still be useful to try skinny instead of 1-convex?}.

@@

iLast <- length(bbhTourFull1)
fProj1 <- bbhTourFull1[[iLast]]
dProj <- as.tibble(bbhM %*% fProj1)
p1 <- ggplot(dProj, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
iLast <- length(bbhTourFull2)
fProj2 <- bbhTourFull2[[iLast]]
dProj <- as.tibble(bbhM %*% fProj2)
p2 <- ggplot(dProj, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
iLast <- length(bbhTourFull3)
fProj3 <- bbhTourFull3[[iLast]]
dProj <- as.tibble(bbhM %*% fProj3)
p3 <- ggplot(dProj, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
grid.arrange(p1, p2, p3, ncol=3) 
```

>>>>

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix analogue to that shown in Figure \ref{fig:neutronStarSPLOM} but where we have removed a subset of less informative parameters to obtain a clear picture. Strong association can be seen between the parameters ra, dec and time, where the first two describe the position of the event in the sky, and time is the merging time (in GPS units). Moreover, the relation between dec and time is not functional, which motivates us to use the TIC index for this example (a drawback is however the run time, this data sample contains 10k points, making optimisation of the MINE index functions very inefficient). In general we observe non-standard distributions for all variables, as well as various patterns of association between sets of variables. These patterns are representative of the likelihood function, since most priors are flat, or build such to capture growth with volume rather than distance. This is still efficient thanks to special sampling techniques that allow jumps in parameter space.

### Data pre-pocessing
We follow the same steps of data pre-processing as described above for the first example. First, since constraints on the masses of two systems observed via gravitational waves are strongly correltated, we again remove the smaller mass m2 from the set of parameters. We then rescale all parameters to the range from 0 to 1.

As a first test we use the full dataset and compare results obtained with different index functions. Here we expect to recover the relation between the correlated variables dec, time and ra. We compare the final view using three different metrics in Figure \ref{fig:bbhGuided}, showing the views with (local?) maximum of the TIC index (left), spines2D index (middle) and for comparison the maximum of 1-convex (right). Indeed the TIC and spline index identify dec, time and ra parameters and show views of the 2-d band that is embedded in three dimensions \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. Notably, the exact views differ based on the index function selected, optimising TIC is showing us the band "from the side". On the other hand, the index based on the scagnostics member convex is too noisy to find any views of interest, and the final  view is essentially equal to the starting plane in this case. All index values for the three final views are listed in Table \ref{tab:indexTableBBH}, where we see that indeed the splines2D index does not change between the first two views, while the TIC index is increased for the first view wrt the second. Finally the value of 1-convex is clearly larger for the first view as compared to the final view, but the optimisation did not succeed for this noisy index function.

@@

iLast <- length(bbhTourFull1)
fProj1 <- bbhTourFull1[[iLast]]
dProj1 <- as.tibble(bbhM %*% fProj1)
p1 <- ggplot(dProj1, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
iLast <- length(bbhTourFull2)
fProj2 <- bbhTourFull2[[iLast]]
dProj2 <- as.tibble(bbhM %*% fProj2)
p2 <- ggplot(dProj2, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
iLast <- length(bbhTourFull3)
fProj3 <- bbhTourFull3[[iLast]]
dProj3 <- as.tibble(bbhM %*% fProj3)
p3 <- ggplot(dProj3, aes(V1,V2)) + geom_point(alpha = 0.02) + theme(aspect.ratio=1)
grid.arrange(p1, p2, p3, ncol=3) 
```

```{r indexTableBBH}
getIdxComp <- function(dProj, n){
  ticI <- mine(dProj$V1, dProj$V2)$TIC
  splineI <- splines2d(dProj$V1, dProj$V2)
  invConvI <- 1-scagnostics(as.matrix(dProj))[,"Convex"]
  return(tibble(TIC=ticI, splines2D=splineI, invConv=invConvI, nView=n))
}
idxT <- tibble(TIC=numeric(),
               splines2D=numeric(),
               invConv=numeric(),
               nView=numeric())
idxT <- rbind(idxT, getIdxComp(dProj1, 1),
      getIdxComp(dProj2, 2),
      getIdxComp(dProj3, 3))
colnames(idxT) <- c("TIC", "splines2D", "1-convex", "Projection")
knitr::kable(idxT,  caption = "Matrix of index values, for all three indices and all three final projections. \\label{tab:indexTableBBH}", digits=2) 
```
